---
title: "Advanced Circumplex Visualization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Circumplex Visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
set.seed(12345)
library(ggplot2)
```

```{r setup}
library(circumplex)
```

## Beyond the built-in plots

The `ssm_plot_circle()`, `ssm_plot_curve()`, `ssm_plot_contrast()`, and
`ssm_plot_trajectory()` functions cover the most common circumplex figures, but
they each produce a finished plot with a fixed set of layers. Sometimes you want
more control: to overlay individual respondents on a group profile, to zoom in on
a band of amplitudes, to restyle the points, or to place several circumplex
panels side by side.

To make that possible, `circumplex` exposes the building blocks that the
built-in plots are themselves made of. These are ordinary
[`ggplot2`](https://ggplot2.tidyverse.org/) components, so you compose them with
`+` and combine them freely with any other `ggplot2` layers, scales, and themes:

- `coord_circumplex()` is the **coordinate system**. It maps the `displacement`
  aesthetic (degrees) onto the angle and the `amplitude` aesthetic onto the
  radius, and it owns the amplitude-to-radius scaling for the whole plot.
- `ggcircumplex()` assembles the empty circular **canvas** --- the coordinate
  system plus the amplitude rings, displacement spokes, and scale labels.
- `geom_ssm_point()` and `geom_ssm_arc()` are the **layers** that place profile
  points and their confidence regions in the circle, taking amplitude and
  displacement directly as aesthetics.
- `theme_circumplex()` is the **theme** the canvas is drawn with, and the rings
  and spokes are ordinary themed panel furniture that respond to further
  theming.
- `scale_x_circumplex()` is a **scale** for the angle axis of linear circumplex
  plots (such as the score-by-angle curve).

This vignette works through each of these and then combines them.

## The circular canvas

`ggcircumplex()` returns a `ggplot2` object containing just the circular
backdrop, with no data drawn on it yet. By default it uses octant scales
labeled by their angular position in degrees:

```{r canvas-default}
ggcircumplex()
```

You can label the scales however you like. Passing a character vector labels the
spokes in the order of the angles:

```{r canvas-labels}
ggcircumplex(octants(), labels = PANO())
```

If you are working with one of the instruments bundled with the package, you can
pass it directly and the scale angles and abbreviations are taken from the
instrument:

```{r canvas-instrument}
ggcircumplex(instrument = csip)
```

Throughout, displacement runs counterclockwise from the right, and the
0/360 degree position is labeled 360.

## The coordinate system

`ggcircumplex()` is a convenience wrapper. Underneath it, the piece that makes a
circumplex plot circular is `coord_circumplex()`, and you can add that to a bare
`ggplot()` yourself when you want to build a figure from scratch:

```{r coord-bare}
results <- ssm_analyze(
  jz2017,
  scales = PANO(),
  measures = c("NARPD", "ASPD")
)
results$results[, c("Label", "a_est", "d_est", "a_lci", "a_uci")]

ggplot(results$results) +
  coord_circumplex(amax = 0.3) +
  geom_ssm_point(aes(amplitude = a_est, displacement = d_est, fill = Label)) +
  theme_circumplex()
```

That is the minimum: a coordinate system and a layer. The spokes fall on
`ggplot2`'s default axis breaks rather than on the circumplex scale angles,
because a bare coordinate system has not been told what the scales are ---
supplying those breaks and labels, along with the theme, is exactly what
`ggcircumplex()` adds on top.

Because the coordinate system owns the amplitude-to-radius mapping, `amax` is
set exactly once per plot and the canvas and the data layers cannot disagree
about what a given radius means. (Earlier versions of the package took an `amax`
argument on each layer; those arguments are now deprecated and ignored, with a
one-time note.) Leaving `amax = NULL` trains it from the data, as
`ssm_plot_circle()` does.

### Moving the center

By default the center of the circle is amplitude 0, so radial distance is
proportional to amplitude and the origin means "no differentiation among the
scales." The `center` argument moves that inner limit, which is useful when
every profile sits in a narrow band of amplitudes and the interesting variation
is squeezed against the rim:

```{r coord-center}
ggplot(results$results) +
  coord_circumplex(amax = 0.28, center = 0.15) +
  scale_x_continuous(breaks = octants(), labels = PANO()) +
  geom_ssm_point(aes(amplitude = a_est, displacement = d_est, fill = Label)) +
  theme_circumplex()
```

This is a zoom, and it changes how the figure should be read. With a nonzero
center, radial distance is no longer proportional to amplitude and the origin no
longer represents zero amplitude, so differences in radius are exaggerated
relative to the default view. The amplitude ring labels still report the true
amplitudes, and they are what the reader should be directed to. Use a nonzero
center to resolve closely spaced profiles, and say so in the caption.

### Moving the amplitude axis

The amplitude (radial) axis and its tick labels are placed automatically in the
widest gap between the displacement spokes, so they never collide with a spoke
label. You can override that with `r_axis_angle`, given as a displacement in
degrees:

```{r coord-r-axis}
ggplot(results$results) +
  coord_circumplex(amax = 0.3, r_axis_angle = 67.5) +
  scale_x_continuous(breaks = octants(), labels = PANO()) +
  geom_ssm_point(aes(amplitude = a_est, displacement = d_est, fill = Label)) +
  theme_circumplex()
```

Note that these examples build the canvas from its parts --- the coordinate
system, an x-scale carrying the spoke breaks and labels, and the theme --- rather
than adding a second coordinate system on top of `ggcircumplex()`, which
`ggplot2` would replace with a message.

## Placing SSM results in the circle

Let's draw the two-measure profile from above on a labeled canvas ourselves,
rather than calling `ssm_plot_circle()`.

`geom_ssm_point()` places a point for each profile at its amplitude (`a_est`)
and displacement (`d_est`), and `geom_ssm_arc()` draws the wedge spanning each
profile's amplitude confidence interval radially and its displacement confidence
interval angularly. Both take the SSM parameters directly as aesthetics and
handle the conversion into circular coordinates internally, including
wrap-around when a displacement interval crosses the 0/360 degree boundary.

```{r results-plot}
ggcircumplex(octants(), labels = PANO(), amax = 0.3) +
  geom_ssm_arc(
    data = results$results,
    mapping = aes(
      amplitude_min = a_lci, amplitude_max = a_uci,
      displacement_min = d_lci, displacement_max = d_uci,
      fill = Label
    ),
    alpha = 0.4, color = NA
  ) +
  geom_ssm_point(
    data = results$results,
    mapping = aes(amplitude = a_est, displacement = d_est, fill = Label)
  )
```

Each arc displays two separate confidence intervals for one profile at once:
its radial extent is the amplitude interval and its angular extent is the
displacement interval. It is a convenient way to show both intervals together,
not a single joint confidence region with its own coverage level, and not a
hypothesis test. The angular extent in particular is a range of plausible
*directions*: because zero degrees is an arbitrary reference direction rather
than a null value, it should not be read as a significance test the way a
confidence interval for a linear parameter (such as elevation) can be.
Displacement is only worth interpreting at all when the amplitude interval is
clearly above zero and the model fits reasonably well (see the "Introduction to
SSM Analysis" vignette and `?ssm_analyze`).

## Restyling the canvas

`theme_circumplex()` is the theme `ggcircumplex()` applies. Because the rings,
spokes, and labels are themed panel furniture rather than drawn geometry, any
further theming reaches them. Adjust the base font size through the theme, and
restyle the gridlines with an ordinary `theme()` call:

```{r theming}
ggcircumplex(octants(), labels = PANO(), amax = 0.3) +
  geom_ssm_point(
    data = results$results,
    mapping = aes(amplitude = a_est, displacement = d_est, fill = Label)
  ) +
  theme_circumplex(base_size = 14) +
  theme(
    panel.grid.major = element_line(color = "steelblue", linetype = "dotted"),
    legend.position = "bottom"
  )
```

## Extending the layers

The `ggplot2` objects behind the layers and the coordinate system ---
`GeomSsmPoint`, `GeomSsmArc`, and `CoordCircumplex` --- are exported, so you can
subclass them the same way you would subclass any `ggplot2` geom. This is the
route to a reusable layer with your own defaults, rather than repeating the same
arguments at every call site.

```{r subclass}
GeomSsmStar <- ggproto("GeomSsmStar", GeomSsmPoint,
  default_aes = modifyList(
    GeomSsmPoint$default_aes,
    list(shape = 23, size = 5, fill = "#D55E00", colour = "black")
  )
)

geom_ssm_star <- function(mapping = NULL, data = NULL, ...) {
  layer(
    geom = GeomSsmStar, mapping = mapping, data = data,
    stat = "identity", position = "identity",
    inherit.aes = FALSE, params = list(...)
  )
}

ggcircumplex(octants(), labels = PANO(), amax = 0.3) +
  geom_ssm_star(
    data = results$results,
    mapping = aes(amplitude = a_est, displacement = d_est)
  )
```

Everything the parent geom does --- the polar placement, the dropping of
profiles with no defined location --- is inherited; only the defaults change.

## Composing custom layers

Because the canvas and geoms are ordinary `ggplot2` objects, you can add
anything else to them. A common request is to show where individual respondents
fall relative to a summary. We can compute each person's own amplitude and
displacement with `ssm_score()` and draw them as a faint cloud behind a
group-level point.

```{r individuals, warning = FALSE}
# Per-person SSM parameters for a subset of the sample. A respondent whose
# scores are flat has no displacement and is returned as NA (with a warning),
# so we keep only the well-defined profiles.
people <- ssm_score(
  jz2017[1:100, ],
  scales = PANO(),
  append = FALSE
)
people <- people[!is.na(people$Disp), ]

# Group-level profile for the same subset
group <- ssm_analyze(jz2017[1:100, ], scales = PANO())

ggcircumplex(octants(), labels = PANO(), amax = 3) +
  geom_ssm_point(
    data = people,
    mapping = aes(amplitude = Ampl, displacement = Disp),
    fill = "grey70", size = 1.5, alpha = 0.6
  ) +
  geom_ssm_point(
    data = group$results,
    mapping = aes(amplitude = a_est, displacement = d_est),
    fill = "#0072B2", size = 4
  )
```

The individual points spread around the circle while the group summary sits near
their center of mass, a picture that none of the built-in functions produce
directly. Any other `ggplot2` layer --- text annotations, additional geoms,
faceting --- can be added the same way.

## Trajectories across occasions

When the same people are measured on the same scales at two or more occasions,
`ssm_analyze_long()` (for long data) or `ssm_analyze(occasions = )` (for wide
data) estimates one SSM profile per occasion, resampling persons so that
within-person dependence across occasions is respected.
`ssm_plot_trajectory()` then draws each SSM parameter against time.

Here is a small simulated three-wave data set whose group profile rotates
counterclockwise across the 0/360 degree boundary, which is the case worth
seeing drawn:

```{r occasions-data}
angles <- octants()
n <- 200
waves <- c(T1 = 330, T2 = 355, T3 = 20) # true displacement at each wave
elevation <- rnorm(n) * 0.5 # per-person offset, stable across waves

long <- do.call(rbind, lapply(names(waves), function(w) {
  # Group signal at this wave, plus the stable person offset and fresh noise
  signal <- 0.6 * cos((angles - waves[[w]]) * pi / 180)
  scores <- matrix(signal, nrow = n, ncol = length(angles), byrow = TRUE) +
    elevation +
    matrix(rnorm(n * length(angles), sd = 0.5), nrow = n)
  out <- as.data.frame(scores)
  names(out) <- PANO()
  out$id <- seq_len(n)
  out$wave <- w
  out
}))

results_long <- ssm_analyze_long(
  long,
  scales = PANO(),
  id = "id",
  occasion = "wave"
)
results_long$results[, c("Occasion", "a_est", "d_est", "d_lci", "d_uci")]
```

```{r occasions-plot, fig.height = 6}
ssm_plot_trajectory(results_long, drop_xy = TRUE)
```

Two things about the displacement panel are worth reading carefully. First, it
is drawn on an *unwrapped* branch: the profile crosses the 0/360 boundary
between the second and third wave, and rather than jumping a full turn the panel
continues past 360, so values outside \[0, 360) are expected there. Second,
occasions appear in the order they were supplied, never in alphabetical order,
so a `T2`/`T10` pair cannot silently swap.

A time point whose amplitude interval is too close to zero for its displacement
to be interpretable is drawn as a hollow point --- a marker of an
interpretability precondition, not a significance test. `drop_xy = TRUE` above
omits the X-value and Y-value panels, leaving elevation, amplitude, and
displacement.

`ssm_plot_trajectory()` also accepts a trajectory table --- a data frame of
`a_est`/`a_lci`/`a_uci` and `d_est`/`d_lci`/`d_uci` triples at numeric time
points --- which is how you plot a *model-based* trajectory evaluated from a
fitted growth model rather than one estimated separately at each wave. That
workflow is the subject of the "Growth Models on SSM Parameters" vignette.

## The angle axis for linear plots

Not every circumplex figure is circular. The score-by-angle curve drawn by
`ssm_plot_curve()` is a linear plot whose x-axis runs through the scale angles.
`scale_x_circumplex()` labels that axis consistently with the circular canvas:
by default with the angle in degrees, or with custom labels or an instrument's
abbreviations.

```{r curve-axis}
curve <- data.frame(
  angle = angles,
  score = 1 + 0.8 * cos((angles - 135) * pi / 180)
)

ggplot(curve, aes(x = angle, y = score)) +
  geom_line() +
  geom_point(size = 2) +
  scale_x_circumplex(angles, labels = PANO()) +
  labs(x = "Scale", y = "Score") +
  theme_bw()
```

Passing the same `labels` (or the same `instrument`) to both `ggcircumplex()`
and `scale_x_circumplex()` guarantees that a circular figure and a linear one
label their scales identically.

## Relationship to the built-in plots

The built-in plotting functions are implemented on exactly these components:
`ssm_plot_circle()` is `ggcircumplex()` plus `geom_ssm_arc()` and
`geom_ssm_point()`, and `ssm_plot_curve()` uses `scale_x_circumplex()` for its
angle axis. So you can always start from a built-in plot and add to it, or
rebuild it from the pieces when you need finer control. Whichever route you
take, the coordinates are computed the same way, so the results line up.

## References

* Gurtman, M. B. (1992). Construct validity of interpersonal personality
  measures: The interpersonal circumplex as a nomological net. _Journal of
  Personality and Social Psychology, 63_(1), 105--118.

* Wright, A. G. C., Pincus, A. L., Conroy, D. E., & Hilsenroth, M. J. (2009).
  Integrating methods to optimize circumplex description and comparison of
  groups. _Journal of Personality Assessment, 91_(4), 311--322.

* Zimmermann, J., & Wright, A. G. C. (2017). Beyond description in interpersonal
  construct validation: Methodological advances in the circumplex Structural
  Summary Approach. _Assessment, 24_(1), 3--23.
</content>
</invoke>
