---
title: "Axes Reliability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Axes Reliability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# This vignette uses lavaan, which circumplex only Suggests. The chunks that
# fit the model are gated on lavaan being installed so the vignette still
# builds without it; chunks that need no lavaan carry an explicit eval = TRUE.
has_lavaan <- requireNamespace("lavaan", quietly = TRUE)
knitr::opts_chunk$set(eval = has_lavaan)
```

```{r setup, eval = TRUE}
library(circumplex)
```

```{r lavaan-note, eval = !has_lavaan, echo = FALSE, results = "asis"}
cat(
  "**Note:** The `lavaan` package is not installed, so the analysis below",
  "is not evaluated. Install it with `install.packages(\"lavaan\")` to run",
  "this vignette."
)
```

## 1. What is axis reliability?

A circumplex instrument places its scales around a circle and summarizes a
person (or a profile of correlations) by their position on two orthogonal
**axes** — here communion (the X axis, at 0°) and agency (the Y axis, at 90°).
Because those axis scores drive everything downstream — a person's projected
location, the displacement and amplitude of a Structural Summary Method
profile — it is worth asking how *reliably* the instrument measures each axis.

`axes_reliability()` answers that question with the estimator of Strack, Jacobs,
and Grosse Holtforth (2013). It fits an item-level measurement model that
decomposes each item's variance into orthogonal pieces — a general factor common
to all items, the two circumplex **axes**, a scale-specificity component, item
error, and, for an instrument administered in blocks, a block-specificity
component — and reads axis reliability off the **axes** component alone. Reliability
is then the Spearman–Brown "list-length" reliability of a composite of that
length built from items that share only their axes variance.

This is a different question from the other reliability-adjacent tools in the
package. `ssm_sem()` disattenuates a *scale-level* SSM profile for measurement
error; `fit_structure()` evaluates whether a correlation matrix has circumplex
*structure* at all. `axes_reliability()` instead reports a single, interpretable
number per axis: how well the instrument measures communion and agency.

## 2. A worked example

The package ships `simulated_items`, a synthetic dataset of 1–7 Likert
responses from 500 respondents on 32 items — four items on each of the eight
octant scales, in the order that `octants()` returns. The items were drawn from
a five-component population (a general factor, two equal axes with axes variance
.18, a shared scale-specificity component of .10, no block specificity — the
instrument is not blockwise, so that component is zero here — and free item
error) that implies an axis reliability of about .78.

`axes_reliability()` needs three things: the data, a map from items to scales,
and the scales' angles. The map is a list with one character vector of item
names per scale, in the same angle order as the angles you pass:

```{r map}
data("simulated_items")

# Four items per octant scale, in octants() order (PA, BC, DE, ..., NO).
items <- split(names(simulated_items), rep(1:8, each = 4))

res <- axes_reliability(simulated_items, items = items, angles = octants())
res
```

If your items belong to one of the package's built-in instruments, you can pass
the `instrument` object instead of `items` and `angles` — it supplies both the
scale angles and the item membership, exactly as `score()` does. (The example
above uses the explicit map because `simulated_items` is not a registered
instrument.)

The header confirms how many complete cases were used, and the per-axis table
reports, for each axis, the effective test length (`item_n`), the Strack axis
`Reliability`, its standard error of measurement (`SEm`), and the
Nunnally–Bernstein reliability (`NB_Reliability`) for comparison. For a balanced
instrument the two axes share one axes-variance estimate and carry equal
`item_n`, so they report the same reliability — expected, not an error.

The recovered reliability (about .77) lands close to the .78 built into the
simulated population, and the axes-variance estimate (below) recovers the
population value of .18.

## 3. Reading the components

`summary()` adds the estimated variance components and the model's global fit:

```{r summary}
summary(res)
```

The **variance components** show the decomposition the reliability rests on:
the `axes` component is the only one that feeds reliability, while `general`,
`scale_specificity`, `block_specificity` (when blocks were supplied), and item
error are isolated from it. This is precisely why
the Nunnally–Bernstein figure printed alongside runs **higher** than the Strack
reliability: N–B charges scale-specificity variance to the axis rather than
isolating it, so it **overestimates** axis reliability whenever scale
specificity is non-trivial (Strack et al., 2013, Figure 3). The gap between the
two numbers is a direct read-out of how much scale-specific variance the simpler
formula would have miscredited to the axes.

## 4. Starting from a published correlation matrix

You do not always have the raw data. A paper may print an item correlation
matrix and nothing else, and that matrix is enough: pass it as `cormat` together
with the sample size it was computed from, in place of `data`.

```{r cormat}
R <- cor(simulated_items)
axes_reliability(
  cormat = R, items = items, angles = octants(), n = nrow(simulated_items)
)
```

The estimates are identical to the raw-data run above — the raw-data path builds
exactly this matrix internally and fits it the same way. `items` selects and
orders the matrix's rows by name, so its own column ordering does not matter,
and it must be symmetric, positive definite, and have a unit diagonal (the model
assumes unit-variance items).

Two things are unavailable here, and both for the same reason: they are
properties of the respondents, not of their correlations. The
Nunnally–Bernstein comparison is reported as `NA` — it needs each scale's alpha
and the axis composite's variance, neither of which a correlation matrix
carries. And `sd = "raw"` is refused, because there are no scale scores to take
an observed SD from; supply the axis SDs numerically if you want SEm on a raw
scale. Both are reported rather than silently omitted, so a matrix-based result
cannot be mistaken for a raw-data one.

## 5. Caveats to keep in mind

Four properties of the method shape how its output should be read.

**The standard errors and global fit are approximate.** Following the paper's
own practice, the model is fit to the item **correlation** matrix as though it
were a covariance matrix. The component point estimates and the reliabilities
are correct, but the component standard errors and the global chi-square are
only approximate (Cudeck, 1989).

How approximate is worth knowing, because it is neither small nor uniform. The
reported standard errors are the ones normal-theory maximum likelihood would
report for a sample **covariance** input, while the estimator actually consumes
a sample **correlation** matrix, whose diagonal cannot vary. For an instrument
whose axes carry a lot of variance the component standard errors therefore
**overstate** sampling variability substantially — by about 40% at an axes
variance of .35 — so an interval built from them errs on the cautious side. For
weak-axes, strong-general instruments the ratio drifts the other way and they
are slightly **understated**. Read the component SEs as order-of-magnitude
guidance rather than calibrated uncertainty; the point estimates,
reliabilities, and SEm are unaffected either way. Read the fit indices as a
rough guide, not an exact test — they are flattered by roughly 4% for the same
reason.

If you check the implied matrix, you will also find that it does
not reproduce the unit diagonal exactly. That too is expected: with the loadings
fixed, the condition a free item error satisfies is the *weighted* diagonal
rather than the raw one, so off-diagonal sampling misfit leaks into the implied
diagonal at roughly the sampling standard error of a correlation.

**Missing data: listwise by default, FIML on request.** The default,
`missing = "listwise"`, uses complete cases only and reports how many there
were. `missing = "fiml"` instead estimates from every respondent who answered
at least one item, and reports the total N with the complete-case count beside
it. Pairwise-deletion correlations are never used on either setting.

The choice is a trade of assumptions, not a free upgrade. FIML requires the
data to be **missing at random** — missingness may depend on values you
observed, but not on the unobserved values themselves — **and multivariate
normal**. Under MCAR, where missingness is unrelated to anything, listwise
deletion is already consistent and merely wasteful, so FIML buys precision
rather than correctness; under MAR, listwise deletion is genuinely biased and
FIML is not. Under FIML the standard errors are observed-information standard
errors on the standardized metric, conditional on the standardization
constants, and are approximate for the same correlation-as-covariance reason as
above. Two results become unavailable, both needing items observed by every
respondent: the Nunnally-Bernstein comparison is `NA` with a stated reason, and
`sd = "raw"` is refused in favour of numeric axis SDs.

Note the provenance, because it differs from the rest of this vignette: Strack
et al. (2013) report no missing-data analyses at all, so nothing about the FIML
path rests on their results. It is certified against this package's own
synthetic oracle, where the true variance components are known by construction.

**A boundary fit returns `NA`, not a clipped value.** If the model estimates an
axes variance outside the interval (0, 1) — at or below zero the axes carry no
variance to be reliable, at or above one they carry all of it — or any negative
variance component, the reliability and SEm are reported as `NA` with a warning
and a boundary flag, rather than a clipped, negative, or missing number. An `NA`
here is a signal that the model did not identify a usable axes-variance
component in your data — not a defect to be worked around.

**A blockwise instrument needs its blocks declared.** Some circumplex
instruments are administered in blocks — items grouped by something other than
their scale — which carries a block-specificity component of its own, reported
as high as 6.7% by Strack et al. (2013, Table 3). Pass `blocks` (a list of item
columns, one element per block, exactly as `items` is a list per scale) and that
component is estimated too: the component table gains a `zeta2` row and
`details$zeta2_fitted` is `TRUE`. Nothing in the bundled instrument objects
records block membership, so the map has to come from you.

The component is only estimable when the blocks say something the model does not
already know. Blocks that coincide with the scales, a single block holding every
item, or one block per item all leave `zeta2` unidentified, and it is dropped
with `details$zeta2_fitted` set to `FALSE` — the same treatment scale
specificity gets on a single-item instrument.

What it costs to *ignore* real blocks depends on their geometry. The general
factor never hands block variance back, so `xi2` is inflated under most layouts
and unchanged under a few — never deflated. The axes variance, the quantity
reliability is actually read from, moves only when block membership carries
information about the *angular distance* between items, beyond what sharing a
scale already tells you.

One case is clean enough to rely on: when **each block draws exactly one item
from every scale**, every within-block pair is a different-scale pair and the
blocks cover every pair of scale positions equally often. Membership then says
nothing about angular distance, and `xi1`, the reliability, and the SEm are
untouched — worth estimating for its own sake, but costless to omit.

Elsewhere, be careful: *"the blocks are spread evenly around the circle" is not
the test.* Blocks pairing diametrically opposite scales are as evenly spread as
blocks get — their angles average to the centre of the circle — and at eight
scales they still pull `xi1` about 9% *below* truth, because every within-block
pair sits exactly half a turn apart, which is very much information about
angular distance. Blocks covering contiguous arcs pull about 12% the other way.
Unless each block takes one item per scale, estimate the component rather than
reasoning about the picture.

Finally, a note on the `SEm`. The standard error of measurement supports a
location interval for a single profile (Strack et al., 2013, use ±1.65·SEm). By
default `axes_reliability()` reports the z-standardized SEm, `sqrt(1 -
reliability)`; pass `sd = "raw"` (or your own axis SDs) to put the SEm on the
raw axis-score scale. Such an interval describes the measurement imprecision of
one profile's axis position; it is not a significance test of that position
against any particular value.

## 6. Wrap-up

`axes_reliability()` gives a compact, per-axis answer to "how reliably does this
instrument measure communion and agency?", isolating the axes variance from the
general and scale-specific components that a simpler reliability formula would
conflate. Use it to characterize a circumplex instrument before leaning on its
axis scores, and read its output with the correlation-as-covariance,
missing-data, and boundary caveats in mind.

The examples above all use the canonical eight octant scales, but nothing in
the model requires them: any **equally spaced** set of angles works, at any
rotation and at any count from four scales upward. Equal spacing is what
matters, and it is required rather than merely preferred — a quasi-circumplex,
whose scales sit at slightly unequal intervals, is refused rather than
approximated, because Strack et al. (2013) excluded such instruments when
validating the model. Three scales are refused for a different reason: at that
count every pair of scales sits the same angular distance apart, so the
general, axes, and scale-specificity variances can no longer be told apart.

Scales may also carry a **single item** each, as Strack et al.'s single-item
circumplex types do. With one item at every position no two items ever share a
scale, so nothing in the data distinguishes scale specificity from item
specificity: the scale-specificity component is not identified, and
`axes_reliability()` drops it from the model rather than estimating a quantity
the data cannot support. The components table then carries three rows instead
of four, and `details$zeta1_fitted` records the drop. An instrument that
*mixes* single-item and multi-item scales still estimates the component — one
multi-item scale supplies the information, and the shared-value restriction
carries it to the remaining scales. One consequence follows for the comparison
figure: coefficient alpha is undefined for a one-item scale, so the
Nunnally–Bernstein reliability is reported as `NA` with its reason whenever any
scale has fewer than two items, exactly as Strack et al. leave that column
blank for such instruments.

One further boundary is worth stating plainly, because the paper crosses it and
this function does not: the model implemented here is **two**-dimensional.
Strack et al. also analyze a *spherical* instrument, SYMLOG, whose items span
three orthogonal dimensions; the figures they publish for it come from that
three-axis model, and no configuration `axes_reliability()` accepts will
reproduce them.

## References

* Cudeck, R. (1989). Analysis of correlation matrices using covariance
  structure models. _Psychological Bulletin, 105_(2), 317–327.

* Strack, S., Jacobs, K. A., & Grosse Holtforth, M. (2013). The reliability of
  circumplex axes. _SAGE Open, 3_(2). https://doi.org/10.1177/2158244013486115
