---
title: "Axes Reliability Caveats"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Axes Reliability Caveats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# This vignette uses lavaan, which circumplex only Suggests. The chunks that
# fit the model are gated on lavaan being installed so the vignette still
# builds without it; chunks that need no lavaan carry an explicit eval = TRUE.
has_lavaan <- requireNamespace("lavaan", quietly = TRUE)
knitr::opts_chunk$set(eval = has_lavaan)
# Output that follows the console width wraps at 77 columns, so with its "#> "
# prefix it fits in 80. Text printed with cat() does not wrap.
options(width = 77)
```

```{r setup, eval = TRUE}
library(circumplex)
```

**Level:** Advanced. Read "Axes Reliability" first.

## 1. Overview

This page continues "Axes Reliability", which runs `axes_reliability()` on the
`simulated_items` data. It states the properties of the method that shape how
its output should be read. Section 2, "The correction for the correlation
metric", explains why the standard errors and the global fit statistics are
corrected. It also says how well the corrected test is calibrated. Section 3, "Missing
data", covers the listwise default and the FIML option. Section 4, "Boundary
fits", says what an `NA` reliability means. Section 5, "Blockwise
instruments", explains when a block-specificity component is estimable and
what ignoring blocks costs. Section 6, "The standard error of measurement",
says what the `SEm` supports. The Wrap-up states which instruments the model
accepts and where it stops, and says that no page follows this one. The
References list the sources cited.

## 2. The correction for the correlation metric

Four properties of the method shape how its output should be read. Sections 2
to 5 take them in turn.

**Both the standard errors and the global fit statistics are corrected for the
correlation metric.**
The model is fit to the item **correlation** matrix as though it were a
covariance matrix, following the paper's own practice. That metric mismatch
leaves the component point estimates and the reliabilities correct. But
anything computed from the input's sampling variability comes out wrong
(Cudeck, 1989).

The component standard errors are **corrected** for it. Normal-theory maximum
likelihood (ML) computes its standard errors for a sample **covariance** input.
This estimator consumes a sample **correlation** matrix, whose diagonal cannot
vary at all. Left uncorrected, that mismatch **overstates** sampling variability
by about 40% for an instrument whose axes carry a lot of variance. It
**understates** sampling variability slightly for instruments with weak axes and
a strong general factor. So the error changes sign across the range of instruments that the
function accepts, which is why it could not be left to a caveat. The reported
SEs are adjusted to the correlation metric and are calibrated uncertainty. They
are typically **smaller** than the standard errors printed in Strack et al.
(2013). Strack et al.'s values come from LISREL (the SEM software they used) and carry the uncorrected approximation. What the
estimator reported before the correction is kept in `details$se_uncorrected`.

The **global fit statistics are corrected too**, for the same mismatch running
the other way. (`chisq` and its `pvalue` test exact fit. `rmsea`, `cfi` and
`srmr` are approximate-fit indices.) Sample correlations vary *less* than the covariances that the
reference chi-square distribution is derived for. So an uncorrected test
statistic comes out too small, and fit is flattered. `chisq`, `pvalue`, `rmsea`
and `cfi` are therefore reported as Satorra–Bentler-type **scaled** values
(Satorra & Bentler, 1994). A scaled value uses the chi-square divided by a
factor computed at the fitted matrix. `cfi` also scales its own baseline model,
the independence model in which all items are uncorrelated.
The factor is recomputed for every fit, not held constant. That is the whole
point, because how much fit was flattered depends on the instrument. `df` and
`srmr` are unchanged, because they are a count of restrictions and a residual
summary, not test statistics with a reference distribution. What the estimator
reported before the scaling is kept in `details$fit_uncorrected`, and the
factors themselves in `details$scaling_factor`.

If you go looking for these numbers in lavaan, match the variant. They are
built with the definitions that lavaan calls `chisq.scaled`, `pvalue.scaled`,
`rmsea.scaled` and `cfi.scaled`, which are the mean-adjusted Satorra–Bentler
forms. They are **not** the `*.robust` ones (`cfi.robust`, `rmsea.robust`),
which apply a different (Brosseau-Liard/Savalei) adjustment and give different
numbers.

The model is estimated with plain ML, and this package applies the
scaling, not lavaan. So `fitMeasures()` on an equivalent fit hands back the
*uncorrected* values under the bare names `chisq`, `pvalue`, `rmsea` and `cfi`.
It also reports no `*.scaled` or `*.robust` measure at all. lavaan supplies
those only for a genuinely scaled estimator such as `"MLM"` or `"MLR"`. When
asked for a measure it does not have, it quietly returns a shorter vector. So a
cross-check against lavaan's bare `cfi` will disagree with `$fit$cfi`. A request
for `cfi.robust` will come back empty rather than disagreeing. Neither outcome is
a defect.

Read the scaled statistic for what it is. The correction makes the test
statistic match its reference distribution *in expectation*, but it is a
calibration, not an exactness guarantee. It will not rescue a model that is
badly wrong. The usual fit-index cutoffs (Hu & Bentler, 1999) are no more or less
appropriate than before. What changed is that the number compared to them no
longer flatters fit in a systematic way.

**How well calibrated is the test, and at what sample size?** The scaling fixes
the metric error, and the χ² test built on it is asymptotically exact. Its
rejection rate approaches the nominal α as N grows relative to the number of
distinct moments, p\* = p(p+1)/2 for p items. The table below gives the rejection rate at
α = .05 from a simulation at one population (eight octant scales, three items
each, axes variance .35).

| p\*/N | 0.50 | 0.25 | 0.12 | 0.06 |
|---|---|---|---|---|
| rejection rate | .092 | .079 | .062 | .054 |

The rate reaches the nominal band by a p\*/N of about 0.06. That is a sweep at a
single population, not a general threshold.

You do not have to work out where your own fit sits on that row. The object
reports both halves of the ratio. `details$n_moments` is p\*, and `details$n` is
the N that the fit was computed at. On the raw-data paths, that N is the count of
rows that the estimator was actually handed. It is smaller than
`details$n_total` wherever cases were dropped. On the correlation-matrix path,
it is the `n` you supplied. Divide the first by the second and read across the
row.

At **N = 600** the test **over-rejects**, measured .06 to .11 at three
populations chosen to bracket the range of instruments that the function
accepts. The uncorrected statistic under-rejects over the same range,
at .02 to .03. It also moves *further* from nominal as N grows. This is because
its error is asymptotic, while the scaled statistic's error is a finite-sample
one that shrinks away. The over-rejection at a fixed N grows with instrument
size (larger `df`) and shrinks with N.

Two practical consequences follow. First, a p-value near whatever threshold you
are using deserves caution at moderate N with many items. But note the
direction: the scaled test **over-flags** misfit rather than flattering it. That
is the safer error, and the opposite of what the uncorrected statistic did.
Second, all of this evidence is **complete-data**. Under `missing = "fiml"` the
scaled statistic is calibrated in mean. But its rejection rate has not been
measured. So none of these rates should be read as applying to that path.

The implied matrix is the matrix that the fitted model predicts for the item
correlations. If
you check it, you will also find that it does not reproduce the unit diagonal
exactly. That too is expected. The loadings (the items' weights on the factors)
are fixed. So the condition that a free item error satisfies is the *weighted*
diagonal, not the raw one. As a result, off-diagonal sampling misfit leaks into the implied diagonal at
roughly the sampling standard error of a correlation.

## 3. Missing data

**Missing data: listwise by default, FIML on request.** The default,
`missing = "listwise"`, uses complete cases only and reports how many there
were. `missing = "fiml"` instead uses full information maximum likelihood
(FIML). It estimates from every respondent who answered at least one item, and
it reports the total N with the complete-case count beside it.
Pairwise-deletion correlations are never used on either setting.

The choice is a trade of assumptions, not a free upgrade. FIML requires the data
to be **missing at random** (MAR) **and multivariate normal**. Under MAR,
missingness may depend on values you observed, but not on the unobserved values
themselves. Under MCAR (missing completely at random), missingness is unrelated
to anything. There, listwise deletion is already consistent and merely
wasteful, so FIML buys precision rather than correctness. Under MAR that is not
MCAR, listwise deletion is genuinely biased and FIML is not.

Under FIML the standard errors are observed-information standard errors on the
standardized metric, conditional on the standardization constants. They carry
the same correlation-metric correction as every other path. The correction is
applied multiplicatively, so the observed information's own account of the
missing data survives it. What the correction does not reach is the uncertainty
in those standardization constants. That uncertainty stays small at mild
missingness but grows with the rate.

At 15% cellwise missingness, the reported SEs average about 7% *below* the
estimator's actual sampling variability. So treat heavy missingness as the case
where they are least trustworthy. Two results become unavailable, because both
need items observed by every respondent. The Nunnally-Bernstein comparison is
`NA` with a stated reason, and `sd = "raw"` is refused in favour of numeric axis
SDs.

Note the provenance, because it differs from the rest of this vignette. Strack
et al. (2013) report no missing-data analyses at all, so nothing about the FIML
path rests on their results. It is certified against this package's own
simulated test data, where the true variance components are known by
construction.

## 4. Boundary fits

**A boundary fit returns `NA`, not a clipped value.** The model can estimate an
axes variance outside the interval (0, 1). At or below zero, the axes carry no
variance to be reliable. At or above one, they carry all of it. In that case, or
if any variance component is negative, the reliability and SEm are reported as
`NA`. The `NA` comes with a warning and a boundary flag, rather than a clipped,
negative, or missing number. An `NA` here is a signal that the model did not
identify a usable axes-variance component in your data. It is not a defect to be
worked around.

## 5. Blockwise instruments

**A blockwise instrument needs its blocks declared.** Some circumplex
instruments are administered in blocks, where items are grouped by something
other than their scale. Such an instrument carries a block-specificity component
of its own, which Strack et al. (2013, Table 3) report as high as 6.7%. Pass
`blocks`, a list of item columns with one element per block, just as `items` is
a list per scale. Then that component is estimated too. The component table
gains a `block_specificity` row (symbol `zeta2`), and `details$zeta2_fitted` is
`TRUE`. Nothing in the bundled instrument objects records block membership, so
the map has to come from you.

The component is only estimable when the blocks say something that the model
does not already know. Blocks that coincide with the scales, a single block
holding every item, or one block per item all leave `zeta2` unidentified. That
is, the data cannot pin down its value. It is then dropped with
`details$zeta2_fitted` set to `FALSE`. This is the same treatment that scale
specificity gets on a single-item instrument (the single-item paragraph of
the Wrap-up).

What it costs to *ignore* real blocks depends on their geometry. The general
factor can take up ignored block variance, but it never gives any back. So
`xi2`, the general-factor variance,
is inflated under most layouts and unchanged under a few, but never deflated.
`xi1`, the axes variance, is the quantity that reliability is actually read
from. It moves only when block membership carries information about the
*angular distance* between items, beyond what sharing a scale already tells you.

One case is clean enough to rely on: **each block draws exactly one item from
every scale**. Then every within-block pair is a different-scale pair, and the
blocks cover every pair of scale positions equally often. Membership says
nothing about angular distance, so `xi1`, the reliability, and the SEm are
untouched. The component is worth estimating for its own sake, but it costs
nothing to omit.

Elsewhere, be careful: *"the blocks are spread evenly around the circle" is not
the test.* Blocks pairing diametrically opposite scales are as evenly spread as
blocks get, because their angles average to the centre of the circle. Yet at
eight scales they still pull `xi1` about 9% *below* truth. The reason is that
every within-block pair sits exactly half a turn apart, and that is very much
information about angular distance. Blocks covering contiguous arcs pull about
12% the other way. Unless each block takes one item per scale, estimate the
component rather than reasoning about the picture.

## 6. The standard error of measurement

Finally, a note on the `SEm`. The standard error of measurement supports a
location interval for a single profile (Strack et al., 2013, use ±1.65·SEm). By
default `axes_reliability()` reports the z-standardized SEm, `sqrt(1 -
reliability)`. Pass `sd = "raw"` (or your own axis SDs) to put the SEm on the
raw axis-score scale. Such an interval describes the measurement imprecision of
one profile's axis position. It is not a significance test of that position
against any particular value.

## Wrap-up

The examples in "Axes Reliability" all use the canonical eight octant scales, but nothing in the
model requires them. Any **equally spaced** set of angles works, at any rotation
and at any count from four scales upward. Equal spacing is what matters, and it
is required rather than merely preferred. A quasi-circumplex, whose scales sit
at slightly unequal intervals, is refused rather than approximated. The reason
is that Strack et al. (2013) excluded such instruments when validating the
model. Three scales are refused for a different reason. At that count every pair
of scales sits the same angular distance apart, so the general, axes, and
scale-specificity variances cannot be told apart.

Scales may also carry a **single item** each, as Strack et al.'s single-item
circumplex types do. With one item at every position, no two items ever share a
scale. So nothing in the data distinguishes scale specificity from item
specificity (the item error component). The scale-specificity component is not identified, and
`axes_reliability()` drops it from the model rather than estimating a quantity
that the data cannot support. The components table then carries three rows
instead of four, and `details$zeta1_fitted` records the drop.

An instrument that *mixes* single-item and multi-item scales still estimates the
component. One multi-item scale supplies the information. The shared-value
restriction (all scales share one scale-specificity variance) carries it to the
remaining scales. One consequence follows for the comparison figure. Coefficient
alpha is undefined for a one-item scale. So the Nunnally–Bernstein reliability
is reported as `NA`, with its reason, whenever any scale has fewer than two
items. Strack et al. leave that column blank for such instruments in the same
way.

One further boundary is worth stating plainly, because the paper crosses it and
this function does not. The model implemented here is **two**-dimensional.
Strack et al. also analyze a *spherical* instrument, SYMLOG, whose items span
three orthogonal dimensions. The figures they publish for it come from that
three-axis model, and no configuration that `axes_reliability()` accepts will
reproduce them.

No page follows this one. The observed-score checks of the same instrument,
the circumplex fit and the interval accuracy, are on "Evaluating Circumplex
Structure".

## References

* Cudeck, R. (1989). Analysis of correlation matrices using covariance
  structure models. _Psychological Bulletin, 105_(2), 317–327.

* Hu, L., & Bentler, P. M. (1999). Cutoff criteria for fit indexes in
  covariance structure analysis: Conventional criteria versus new
  alternatives. _Structural Equation Modeling, 6_(1), 1–55.

* Satorra, A., & Bentler, P. M. (1994). Corrections to test statistics and
  standard errors in covariance structure analysis. In _Latent variables
  analysis: Applications for developmental research_ (pp. 399–419).

* Strack, S., Jacobs, K. A., & Grosse Holtforth, M. (2013). The reliability of
  circumplex axes. _SAGE Open, 3_(2). https://doi.org/10.1177/2158244013486115
