---
title: "Evaluating Circumplex Structure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Evaluating Circumplex Structure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(circumplex)
```

## 1. Two questions to ask before interpreting an SSM analysis

The Structural Summary Method (SSM) condenses a circumplex profile into a
few interpretable parameters: elevation, amplitude, displacement, and fit.
That condensation rests on two assumptions that are testable but often left
untested:

1. **Does the instrument actually have circumplex structure?** The SSM
   locates a profile on a circle whose positions are given by the scales'
   theoretical angles. If the scales do not really form a circumplex — or
   sit at very different angles than the theory says — the summary is
   distorted before any profile is computed.

2. **Are the confidence intervals trustworthy at your sample size and
   profile?** Zimmermann and Wright (2017) showed that the accuracy of
   bootstrap SSM intervals depends on the sample size, the instrument's
   correlation structure, and — for amplitude and displacement — on how
   differentiated the profile really is in the population. Intervals that
   are accurate for one construct in your table can be inaccurate for the
   construct printed directly below it.

This vignette shows how to answer both questions with `cpm_fit()` (question
1) and `ssm_ci_accuracy()` (question 2), using the `jz2017` dataset — the
same sample of 1,166 undergraduates analyzed in Zimmermann and Wright
(2017, Study 5), with octant scores on the IIP-SC and personality disorder
(PD) scales from the PDQ-4+.

## 2. Does the instrument fit a circumplex?

### Fitting the circular process model

`cpm_fit()` estimates Browne's (1992) circular process model (CPM), the
confirmatory model behind the CIRCUM and CircE programs. Each scale gets an
estimated angle on the circle and a *communality index* $\zeta$ (the
correlation between the scale and the common circumplex "factor"; its
square, $\zeta^2$, is the scale's *communality* — the share of its
variance the circumplex explains). The
model also yields the usual covariance-structure fit indices.

On the raw-data path the default confidence intervals are bootstrapped. We
lower `boots` from its default of 2000 to keep this vignette quick to
build; in practice, use the default.

```{r cpm}
data("jz2017")
set.seed(12345)
cpm <- cpm_fit(jz2017, scales = PANO(), angles = octants(), boots = 500)
summary(cpm)
```

Two parts of this output matter most for evaluating structure:

- **The estimated angles.** Compare them with the theoretical angles: the
  IIP-SC octants land near — but not exactly on — their theoretical
  positions, so the circumplex ordering is preserved while the *spacing* is
  not perfectly equal. This is the typical result for well-validated
  circumplex instruments: minor deviations from perfect structure are
  common and have little practical impact on SSM profiles (Gurtman &
  Pincus, 2000).
- **The communality indices.** Scales with low $\zeta$ are poorly
  described by the circle; a profile peak at such a scale's angle means
  less than the same peak at a well-explained scale's angle.

The estimated configuration is easiest to read on the circle itself: each
scale is drawn at its estimated angle, at a radius equal to its communality
($\zeta^2$), with a joint confidence region where one is estimable.

```{r cpm_plot, fig.width = 7.5, fig.height = 4.5, out.width = "100%"}
plot(cpm)
```

### Reading the fit indices

The `summary()` output reports the standard indices, and the standard
benchmarks apply: RMSEA at or below about .08 suggests adequate
approximate fit and above about .10 poor fit (Browne & Cudeck, 1993); SRMR
at or below about .08 suggests acceptable average residuals (Hu & Bentler,
1999); CFI and TLI near .95 or above suggest good comparative fit (Hu &
Bentler, 1999). Treat these as conventions from the broader
covariance-structure literature, not circumplex-specific laws.

Two circumplex-specific cautions, both from this package's own validation
simulations:

- **Boundary solutions are common at realistic sample sizes.** When octant
  scales share a strong general factor (as interpersonal problem scales
  do), fitted solutions frequently sit at or near a parameter boundary — a
  communality estimate at 1 (a "Heywood" case), or a harmonic weight at 0.
  `cpm_fit()` flags these in its diagnostics. They are usually a property
  of the estimator meeting real data at finite $n$, not a data-entry error,
  but they matter for inference (next point).
- **Do not lean hard on the chi-square p value.** In simulations at
  field-typical sample sizes with octant-like population structures, the
  test statistic did not follow its nominal chi-square reference
  distribution, even for correctly specified models. Read the chi-square
  descriptively (bigger is worse), and prefer RMSEA/SRMR/CFI for judging
  approximate fit.

For the same reasons, when the intervals are analytic (Wald) — the only
option on the correlation-matrix input path — `summary()` prints a caution
below $N = 2000$, and up to very large $N$ when a boundary marker is
present: in validation, analytic intervals mis-covered in exactly those
regimes. Prefer the bootstrap (the raw-data default) when you have raw
data.

### Comparing model variants

Constrained variants make the structural question sharp: is the instrument
consistent with *equally spaced* scales? With *equal* communalities? The
correlation-matrix path is convenient for this comparison (point estimates
and fit indices are deterministic, so no bootstrap is needed):

```{r variants}
R <- cpm$matrices$R # the sample correlation matrix stored by cpm_fit()
fits <- lapply(
  c("quasi-circumplex", "equal-communality", "circulant"),
  function(mod) {
    cpm_fit(
      cormat = R, scales = PANO(), angles = octants(),
      n = nrow(jz2017), model = mod
    )
  }
)
data.frame(
  model = vapply(fits, function(f) f$details$model, character(1)),
  df = vapply(fits, function(f) f$fit$df, numeric(1)),
  rmsea = round(vapply(fits, function(f) f$fit$rmsea, numeric(1)), 3),
  srmr = round(vapply(fits, function(f) f$fit$srmr, numeric(1)), 3),
  cfi = round(vapply(fits, function(f) f$fit$cfi, numeric(1)), 3),
  tli = round(vapply(fits, function(f) f$fit$tli, numeric(1)), 3)
)
```

The pattern reproduces what Zimmermann and Wright (2017, p. 14) reported
for these data with CircE: the fully constrained model (equal spacing *and*
equal communality — the `"circulant"` variant) fits poorly, while relaxing
the equal-spacing constraint improves fit to the edge of the conventional
benchmarks (marginal by RMSEA, acceptable by SRMR and CFI); they reported
CFI = .824, TLI = .795, RMSEA = .169 for the former and CFI = .958, TLI =
.931, RMSEA = .098 for the latter. Do not expect our indices to match theirs to
the digit: CIRCUM/CircE fit a covariance version of the model with free
scaling constants, whereas `cpm_fit()` fits the correlation structure
directly, and the two differ slightly at finite sample sizes. The
*conclusion* — ordered octants with unequal spacing, adequate approximate
fit once equal spacing is not forced — is the same.

A poor CPM fit does not make SSM output uncomputable; it makes the summary
less meaningful, because the profile is being referred to scale positions
the data contradict. If the ordering itself fails (scales out of sequence
around the circle), SSM parameters should not be interpreted.

## 3. Can you trust your confidence intervals?

### What Zimmermann & Wright (2017) found

Zimmermann and Wright's simulation studies (their Studies 1–4) evaluated
the accuracy of 95% percentile bootstrap intervals for SSM parameters —
the same interval type `ssm_analyze()` reports — judging an interval
accurate when its empirical coverage stayed within Bradley's (1978)
liberal band of 92.5% to 97.5%. Their headline results:

| Parameter | Point estimate | 95% CI accurate when... |
|---|---|---|
| Elevation ($e$) | essentially unbiased | $n \ge 50$ |
| X value / affiliation | essentially unbiased | $n \ge 50$ |
| Y value / dominance | essentially unbiased | $n \ge 50$ |
| Amplitude ($a$) | **biased upward**, strongly so when population amplitude is small | $n \ge 75$ (general-factor instrument) or $n \ge 150$ (no general factor), *given* population amplitude $\ge .10$ |
| Displacement ($\delta$) | unbiased but imprecise at low amplitude | $n \ge 100$ (general factor) or $n > 200$ (no general factor), *given* population amplitude $\ge .10$ |
| Fit ($R^2$) | biased downward | population $R^2 < .9$ only; unsuited near 1 |

*(Transcribed from Zimmermann & Wright, 2017, Studies 1–2, pp. 6–11.)*

Three implications are worth internalizing:

- **Amplitude overestimates differentiation.** Across their conditions the
  relative bias averaged 15.5% and reached 135.8%; with $n = 50$ and no
  general factor, the *expected* sample amplitude was about .15 when the
  population amplitude was exactly zero (their pp. 6–7). An amplitude of
  .15 is commonly read as "marked"; at small $n$ it can be pure sampling
  artifact.
- **Displacement is only as stable as amplitude is large.** The standard
  error of the displacement grows as amplitude shrinks (e.g., roughly 50
  degrees at $n = 100$ for a weakly differentiated profile; their p. 8).
  A displacement estimate from a flat profile has no meaningful direction.
- **The thresholds above are grid minima, not guarantees.** They come from
  specific instruments, specific sample sizes, and population amplitudes
  of .10 or larger. Their Study 3 quantified the frontier more finely: the
  minimum population affiliation/dominance component needed for accurate
  amplitude and displacement intervals is approximately
  $2.95 \cdot f_a \cdot n^{-0.587}$, where $f_a$ is an instrument
  constant (their Eq. 3; ≈ .55 for the IIP-C, .63 for the IIP-SC, .85 for
  the IAS). At $n = 100$ with an IIP-C-like instrument that is a
  population component of about .11 — larger than many real construct
  profiles.

### Simulating coverage for your own analysis

Published thresholds cover a coarse grid of conditions, and your analysis
— your instrument, your $n$, your group sizes, your profile, your number
of bootstrap resamples — usually falls between or outside the grid points.
`ssm_ci_accuracy()` answers the question directly at your configuration:
it builds a population whose structure matches your fitted estimates
(through the CPM by default), simulates many datasets of your exact sample
size from it, replays your own CI procedure on each, and reports how often
the intervals covered the known population values.

To see it catch a real problem, we analyze two PD scales at a modest
sample size: paranoid PD (a well-differentiated interpersonal profile) and
obsessive–compulsive PD, whose profile Zimmermann and Wright's Table 4
showed to be nearly flat (amplitude .012 at full sample size). We use the
first 250 participants — a sample size many construct-validation studies
would consider respectable.

```{r accuracy}
set.seed(23456)
res <- ssm_analyze(
  jz2017[1:250, ],
  scales = PANO(),
  angles = octants(),
  measures = c("PARPD", "OCPD"),
  boots = 500 # reduced from the default 2000 to keep the vignette quick;
  # the diagnostic below replays whatever procedure this object used
)
summary(res)
```

Note the guardrail in the printed output: when a profile's amplitude
interval includes zero, `ssm_analyze()` marks the displacement as
uninterpretable rather than certifying its direction. The diagnostic below
measures, among other things, how well that certification rule performs at
your configuration.

We again use reduced settings (`reps = 200`, a short amplitude ladder, and
the smaller `boots` above) to keep the vignette fast; the defaults
(`reps = 1000`, four ladder rungs, on an object built with the default
`boots = 2000`) give a Monte Carlo standard error under one percentage
point for coverage near the nominal level and are recommended in practice.
The `parallel`/`ncpus` arguments speed up real runs without changing
results for a given seed.

```{r accuracy_run}
set.seed(34567)
acc <- ssm_ci_accuracy(res, reps = 200, amplitude_factors = c(1, 0.5, 0))
summary(acc)
```

How to read this output:

- **The coverage table** reports, for each parameter, how often your CI
  procedure covered the truth in the simulated replications, at the
  amplitude the profile was estimated to have (`Condition = 1`) and at
  fractions of it. Compare coverage against the nominal level using the
  Wilson interval and the Bradley band, which is what the verdict does for
  you.
- **The amplitude ladder** (`Condition` column) matters because your
  estimated amplitude is biased upward — the population that generated
  your data plausibly has *less* differentiation than your estimate. The
  rungs below 1 show what happens to coverage in that direction; the 0
  rung shows the degenerate flat-profile case, where displacement has no
  true value at all and the amplitude interval cannot cover the boundary
  truth (its printed zero coverage is structural, flagged in the table).
  When a profile's estimated amplitude is smaller than half its own CI
  width — as for obsessive–compulsive PD here — the diagnostic adds one
  more rung at the *certification margin* (a scaling factor above 1),
  the amplitude at which the certification threshold would sit exactly at
  the observed half-width.
- **The guardrail table** reports how often profiles were *certified*
  (amplitude CI excluding zero) at each rung. At the 0 rung, any
  certification is a false certification: the summary prints a caution
  when the false-certification rate is materially above the rate a user
  would expect from the interval level. This is a measured property of the
  shipped display rule, not a hypothesis test.
- **The verdict** classifies elevation, amplitude, and
  certification-conditional displacement coverage at your as-estimated
  amplitude, and the plain-language paragraph states the overall
  conclusion per profile.

For a visual summary across the ladder:

```{r accuracy_plot, fig.width = 7.5, fig.height = 5, out.width = "100%"}
plot(acc)
```

The two profiles tell usefully different stories. For paranoid PD the
intervals themselves behave: nothing in its coverage table clearly leaves
the Bradley band, at the as-estimated amplitude or with the population
amplitude halved. For
obsessive–compulsive PD, amplitude and certified-displacement coverage
fall well below the Bradley band already at the profile's own estimated
amplitude — its intervals under-cover, in the direction of missing below
the truth.

Yet *both* verdicts lead with a caution, and the reason is the guardrail
line: at this configuration ($n = 250$, eight octant correlations), a
profile whose true amplitude is exactly zero would still be certified
essentially every time. The "amplitude CI excludes zero" rule offers
almost no protection here — certification is not, by itself, evidence of
real differentiation. What carries the reassurance for paranoid PD is the
coverage evidence across the ladder, not the certificate. This is the
same phenomenon Zimmermann and Wright flagged from the estimation side:
at modest $n$, a sample amplitude large enough to clear zero is the
*expected* outcome even from a flat population profile.

One reading note: at reduced `reps`, classes tend to print as
`borderline` — the Wilson interval around the estimated coverage is too
wide to place it clearly inside or outside the Bradley band. That is the
diagnostic being honest about its own Monte Carlo error; the default
`reps = 1000` sharpens such classifications into `adequate` or
`inadequate`.

Two settings worth knowing: `structure = "observed"` rebuilds the
population from the pooled observed correlations instead of the CPM — if
the two structures yield different verdicts, that disagreement is itself
informative (structure uncertainty is material for your data); and the
embedded CPM fit is returned as `acc$cpm` for inspection with the tools
from Section 2.

The diagnostic itself was validated against Zimmermann and Wright's
published results: configured to their transcribed simulation conditions,
it reproduces their accuracy classifications — coverage inside the Bradley
band at conditions they found accurate, and the same one-sided
under-coverage at conditions they found inaccurate.

### When to trust SSM parameters

Putting Sections 2 and 3 together into a checklist:

1. **Structure first.** Fit `cpm_fit()` to your instrument in your sample.
   If the octant ordering fails or communalities are very low, stop; SSM
   positions mean little.
2. **Elevation is the robust parameter.** Essentially unbiased, and its
   intervals were accurate from $n \ge 50$ in every published and
   package-run condition.
   (On the correlation path, elevation is also the parameter ipsatizing
   destroys — see Section 4.)
3. **Treat amplitude as optimistic.** It cannot go below zero, so it
   overshoots when true differentiation is weak. Before interpreting a
   "marked" amplitude at modest $n$, run `ssm_ci_accuracy()` and look at
   the sub-1 ladder rungs.
4. **Only interpret displacement when amplitude is credible.** The package
   already withholds the displacement interval when the amplitude interval
   includes zero; the diagnostic tells you how reliable that certification
   is at your configuration.
5. **Do not build claims on the fit parameter's interval.** Zimmermann and
   Wright found bootstrap $R^2$ intervals accurate only when population
   fit was mediocre — near-perfect prototypicality cannot be bracketed
   from below. Report $R^2$ descriptively.
6. **Remember what the diagnostic conditions on.** It asks "would my CI
   procedure work in a population *like my estimates*", under multivariate
   normality, with complete data. It is a strong check, not a certificate.

## 4. Ipsatization and what it costs

Ipsatizing — subtracting each respondent's own mean across the octant
scales from each of their scale scores (`ipsatize()`) — is a common
preprocessing step in circumplex work, used to remove individual
differences in overall endorsement before examining profile shape.

For SSM analyses its main cost is simple to state: **ipsatizing discards
elevation.** After row-centering, every respondent's octant scores sum to
zero, so a group's mean profile has zero mean by construction, and the
covariances between the ipsatized scales and any external measure sum to
exactly zero — forcing the mean correlation toward zero regardless of how
strongly the construct relates to the instrument's general factor
(Zimmermann & Wright, 2017, p. 4).

```{r ipsatize}
set.seed(45678)
res_raw <- ssm_analyze(
  jz2017,
  scales = PANO(),
  angles = octants(),
  measures = "PARPD",
  boots = 100
)
jz_ips <- ipsatize(jz2017, items = PANO())
set.seed(45678)
res_ips <- ssm_analyze(
  jz_ips,
  scales = paste0(PANO(), "_i"),
  angles = octants(),
  measures = "PARPD",
  boots = 100
)
comparison <- rbind(
  res_raw$results[, c("e_est", "a_est", "d_est")],
  res_ips$results[, c("e_est", "a_est", "d_est")]
)
rownames(comparison) <- c("raw", "ipsatized")
round(comparison, 3)
```

The raw-score elevation — which matches the value Zimmermann and Wright
report for this scale in their Table 4 — collapses to near zero after
ipsatizing. The displacement stays close to its raw-score value, and the
amplitude stays broadly similar — though not identical, because ipsatizing
also changes the scales' variances and intercorrelations, so shape
parameters shift somewhat too. Guidance:

- If elevation carries meaning in your application (for interpersonal
  problems it indexes association with generalized interpersonal
  distress), analyze raw scores and let the SSM separate elevation from
  shape — that separation is exactly what the model is for.
- If you receive data that were already ipsatized, do not interpret the
  elevation row, and say so in the write-up; amplitude and displacement
  remain interpretable.
- Do not describe an ipsatized profile's near-zero elevation as evidence
  that a construct is "not generally interpersonal" — the preprocessing
  made that value uninformative.

## 5. Wrap-up

Evaluating circumplex structure has two layers: whether the instrument
behaves like a circumplex in your sample (`cpm_fit()`), and whether the
inferential machinery of the SSM can be trusted at your sample size and
profile (`ssm_ci_accuracy()`). Both are one function call, and both change
what you should claim more often than users expect — amplitude and
displacement intervals, in particular, earn their trust only when the
profile is genuinely differentiated relative to the precision your sample
size affords.

## References

* Bradley, J. V. (1978). Robustness? _British Journal of Mathematical and
  Statistical Psychology, 31_(2), 144–152.

* Browne, M. W. (1992). Circumplex models for correlation matrices.
  _Psychometrika, 57_(4), 469–497.

* Browne, M. W., & Cudeck, R. (1993). Alternative ways of assessing model
  fit. In K. A. Bollen & J. S. Long (Eds.), _Testing structural equation
  models_ (pp. 136–162). Newbury Park, CA: Sage.

* Grassi, M., Luccio, R., & Di Blas, L. (2010). CircE: An R implementation
  of Browne's circular stochastic process model. _Behavior Research
  Methods, 42_(1), 55–73.

* Gurtman, M. B., & Pincus, A. L. (2000). Interpersonal Adjective Scales:
  Confirmation of circumplex structure from multiple perspectives.
  _Personality and Social Psychology Bulletin, 26_(3), 374–384.

* Hu, L., & Bentler, P. M. (1999). Cutoff criteria for fit indexes in
  covariance structure analysis: Conventional criteria versus new
  alternatives. _Structural Equation Modeling, 6_(1), 1–55.

* Zimmermann, J., & Wright, A. G. C. (2017). Beyond description in
  interpersonal construct validation: Methodological advances in the
  circumplex Structural Summary Approach. _Assessment, 24_(1), 3–23.
