---
title: "Introduction to SSM Analysis"
author: "Jeffrey M. Girard"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to SSM Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(circumplex)
library(ggplot2)
library(ggforce)
library(tibble)
library(kableExtra)
```

## 1. Background and Motivation

### Circumplex models, scales, and data
Circumplex models are popular within many areas of psychology because they offer
a parsimonious account of complex psychological domains, such as emotion and
interpersonal functioning. This parsimony is achieved by understanding phenomena 
in a domain as being a "blend" of two primary dimensions. For instance, 
circumplex models of emotion typically represent affective phenomena as a blend
of *valence* (pleasantness versus unpleasantness) and *arousal* (activity versus
passivity), whereas circumplex models of interpersonal functioning typically 
represent interpersonal phenomena as a blend of *communion* (affiliation
versus separation) and *agency* (dominance versus submissiveness). These models
are often depicted as circles around the intersection of the two dimensions (see
figure). Any given phenomenon can be located within this circular space through
reference to the two underlying dimensions (e.g., anger is a blend of
unpleasantness and activity).

Circumplex scales contain multiple subscales that attempt to measure different
blends of the two primary dimensions (i.e., different parts of the circle).
Although there have historically been circumplex scales with as many as sixteen
subscales, it has become most common to use eight subscales: one for each "pole"
of the two primary dimensions and one for each "quadrant" that combines the two
dimensions. In order for a set of subscales to be considered circumplex, they
must exhibit certain properties. Circumplex fit analyses can be used to quantify
these properties.

Circumplex data is composed of scores on a set of circumplex scales for one or
more participants (e.g., persons or organizations). Such data is usually
collected via self-report, informant-report, or observational ratings in order
to locate psychological phenomena within the circular space of the circumplex
model. For example, a therapist might want to understand the interpersonal
problems encountered by an individual patient, a social psychologist might want
to understand the emotional experiences of a group of participants during an
experiment, and a personality psychologist might want to understand what kind
of interpersonal behaviors are associated with a trait (e.g., extraversion).

<!-- ![](image) -->

### The Structural Summary Method
The Structural Summary Method (SSM) is a technique for analyzing circumplex data
that offers practical and interpretive benefits over alternative techniques. 

### Understanding the SSM parameters

## 2. Example data: jz2017
To illustrate the SSM functions, we will use the example dataset `jz2017`, which
was provided by Zimmermann & Wright (2017) and reformatted for this package.
This dataset includes self-report data from 1166 undergraduate students.
Students completed a circumplex measure of interpersonal problems with eight 
subscales (PA, BC, DE, FG, HI, JK, LM, and NO) and a measure of personality
disorder symptoms with ten subscales (PARPD, SCZPD, SZTPD, ASPD, BORPD, HISPD,
NARPD, AVPD, DPNPD, and OCPD). More information about these variables can be
accessed using the `?jz2017` command in R.

```{r}
data("jz2017")
print(jz2017)
```

The circumplex scales in `jz2017` come from the Inventory of Interpersonal
Problems - Short Circumplex (IIP-SC). These scales can be arranged into the
following circular model, which is organized around the two primary dimensions
of agency (y-axis) and communion (x-axis). Note that the two-letter scale 
abbreviations and angular values are based in convention. A high score on PA
indicates that one has interpersonal problems related to being "domineering" or
too high on agency, whereas a high score on DE indicates problems related to 
being "cold" or too low on communion. Scales that are not directly on the y-axis
or x-axis (i.e., BC, FG, JK, and NO) represent blends of agency and communion.

```{r, echo = FALSE, fig.width = 7.2, fig.height = 4}
angles <- c(90, 135, 180, 225, 270, 315, 360, 45)
flabel <- c(
  "Domineering", 
  "Vindictive", 
  "Cold", 
  "Socially\nAvoidant", 
  "Nonassertive", 
  "Easily\nExploited", 
  "Overly\nNurturant", 
  "Intrusive")
alabel <- c("PA", "BC", "DE", "FG", "HI", "JK", "LM", "NO")

# Create plot ------------------------------------------------------------------

ggplot() +
  # Require plot to be square and remove default styling
  coord_fixed() +
  theme_void() +
  # Expand both axes multiplicatively to fit the labels
  scale_x_continuous(expand = c(0.30, 0)) +
  scale_y_continuous(expand = c(0.10, 0)) +
  # Draw line segments corresponding to the octants
  geom_segment(
    aes(
      x = 0,
      y = 0,
      xend = 5 * cos(angles * pi / 180),
      yend = 5 * sin(angles * pi / 180)
    ),
    color = "gray60",
    size = 0.5
  ) +
  # Draw inner labels for the octant abbreviations
  geom_label(
    aes(
      x = 3.75 * cos(angles * pi / 180),
      y = 3.75 * sin(angles * pi / 180),
      label = alabel
    ),
    size = 5,
    color = "gray40",
    label.size = NA,
    hjust = "center",
    vjust = "center"
  ) +
  # Draw inner labels for the octant angles
  geom_label(
    aes(
      x = 2 * cos(angles * pi / 180),
      y = 2 * sin(angles * pi / 180),
      label = sprintf("%d\u00B0", angles)
    ),
    size = 4,
    color = "gray50",
    label.size = NA,
    hjust = "center",
    vjust = "center"
  ) +
  # Draw the circle
  geom_circle(aes(x0 = 0, y0 = 0, r = 5),
    color = "gray50", size = 1.5
  ) +
  # Draw outer labels for the octant names
  geom_label(
    aes(
      x = 5.1 * cos(angles * pi / 180),
      y = 5.1 * sin(angles * pi / 180),
      label = flabel
    ),
    size = 5,
    color = "black",
    label.size = NA,
    hjust = "outward",
    vjust = "outward"
  )
```

## 3. Mean-based SSM Analysis

### Conducting SSM for a group's mean scores
To begin, let's say that we want to use the SSM to describe the interpersonal 
problems of the average individual in the entire dataset. We can use the
`ssm_analyze()` function to perform this analysis. The first argument is always
`.data`, a data frame containing all the variables to be analyzed. The second
argument is always `scales` and specifies where in `.data` the circumplex scales
are (either in terms of their variable names or their column numbers). The third
argument is always `angles` and specifies the angular displacements of each of
the  circumplex scales included in `scales`. Note that the `scales` and `angles`
arguments need to be vectors, so if multiple names/angles are provided, they 
need to be combined into a vector using the `c()` function. The order and length
of the `scales` and `angles` vectors also needs to be the same.

```{r}
results <- ssm_analyze(.data = jz2017,
  scales = c(PA, BC, DE, FG, HI, JK, LM, NO),
  angles = c(90, 135, 180, 225, 270, 315, 360, 45))
```

The output of the function has been saved in the `results` variable, which we 
can examine in detail using the `summary()` function. This will output the call
we made to create the output, as well as some of the default options that we
didn't bother changing and, most importantly, the estimated SSM parameter values
with bootstrapped confidence intervals.

```{r}
summary(results)
```

That was pretty easy! We can now write up these results. However, the
`circumplex` package has some features that can make what we just did even 
easier. First, because the first three arguments of the `ssm_analyze()`
function are always the same, we can omit their names. Second, because we 
organized the `jz2017` data frame to have the circumplex scale variables
adjacent and in order from PA to NO, we can simplify their specification by 
using the `PA:NO` shortcut. Finally, because the use of octant scales is so 
common, the `circumplex` package comes with a convenience function for outputing
their angular displacements: `octants()`. Note how, even when using these
shortcuts, the results are the same except for minor stochastic differences in
the confidence intervals due to the randomness inherent to bootstrapping. (To
get the exact same results, we could use the `set.seed()` function to control
the random number generator in R.)

```{r}
results2 <- ssm_analyze(jz2017, PA:NO, octants())
summary(results2)
```

### Visualizing the results with a table and figure
Next, we can produce a table to display our results. With only a
single set of parameters, this table is probably overkill, but in future
analyses we will see how this function saves a lot of time and effort. To create
the table, simply pass the `results` object to the `ssm_table()` function.

```r
ssm_table(results)
```

```{r, echo = FALSE}
ssm_table(results2, render = FALSE) %>%
  kable() %>% 
  kable_styling()
```

On this screen, the numbers don't fit perfectly, so let's free up some
horizontal space by omitting the X-Value and Y-Value columns. (These columns are
useful for testing hypotheses about primary dimensions, but aren't really needed
here since our main goal is just to understand the SSM procedure.)

```r
ssm_table(results2, xy = FALSE)
```

```{r, echo = FALSE}
ssm_table(results2, xy = FALSE, render = FALSE) %>%
  kable() %>% 
  kable_styling()
```

Next, let's leverage the fact that we are working within a circumplex space by
creating a nice-looking circular plot by mapping the amplitude parameter to the
points' distance from the center of the circle and the displacement parameter
to the points' rotation from due-east (as is conventional). This, again, is as 
simple as passing the `results` object to the `ssm_plot()` function.

```{r, fig.width = 7.2, fig.height = 4}
ssm_plot(results2)
```

Very cool! We now have a table and figure that can be included in a write-up,
and the whole process required less than five lines of code.

## 4. Correlation-based SSM Analysis

### Conducting SSM for a group's correlations with an external measure
Next, let's say that we are interested in analyzing not the mean scores on the
circumplex scales but rather their correlations with an external measure. This
is sometimes referred to as "projecting" that external measure into the
circumplex space. As an example, let's project the NARPD variable, which
captures symptoms of narcissistic personality disorder, into the circumplex
space defined by the IIP-SC. Based on theory and previous findings, we can
expect this measure to be associated with some general interpersonal distress
and a style that is generally high in agency.

To conduct this analysis, we can start with the syntax from the mean-based
analysis. All SSM analyses use the `ssm_analyze()` and the data, scales, and
angles are the same as before. However, we also need to let the function know
that we want to analyze correlations with NARPD as opposed to scale means. To
do this, we add an additional argument `measures`.

```{r}
results3 <- ssm_analyze(jz2017, PA:NO, octants(), measures = NARPD)
summary(results3)
```

Note that this output looks very similar to the mean-based output except that
the statistical basis is now correlation scores instead of mean scores and
instead of saying "Profile [All]" it now says "Profile [NARPD]".

### Visualizing the results with a table and figure
We can also create a similar table and figure using the exact same syntax as
before. The `ssm_table()` and `ssm_plot()` functions are smart enough to know
whether the results are mean-based or correlation-based and will work in both
cases.

```r
ssm_table(results3, xy = FALSE)
```

```{r, echo = FALSE}
ssm_table(results3, xy = FALSE, render = FALSE) %>%
  kable() %>% 
  kable_styling()
```

From the table, we can see that the model fit is good (>.80) and that all three
SSM parameters are significantly different from zero, i.e., their confidence
intervals do not include zero. Furthermore, the confidence intervals for the 
elevation and amplitude parameters are greater than or equal to 0.15, which can
be interpreted as being "marked." So, consistent with our hypotheses, NARPD was
associated with marked general interpersonal distress (elevation) and was
markedly distinctive in its profile (amplitude). The displacement parameter was
somewhere between 100 and 120 degrees; to interpret this we would need to either
consult the mapping between scales and angles or plot the results.

```{r, fig.width = 7.2, fig.height = 4}
ssm_plot(results3)
```

From this figure, it is very easy to see that, consistent with our hypotheses,
the displacement for NARPD was associated with high agency and was somewhere
between the "domineering" and "vindictive" octants.

## 5. Wrap-up
In this vignette, we learned about circumplex models, scales, and data as well
as the Structural Summary Method (SSM) for analyzing such data. We learned about
the `circumplex` package and how to use the `ssm_analyze()` function to generate
SSM results for a single group's mean scores and for correlations with a single
external measure. We learned several shortcuts for making calls to this function
easier and then explored the basics of SSM visualization by creating simple
tables and circular plots. In the next vignette, "Intermediate SSM Analysis",
we will build upon this knowledge to learn how to (1) generalize our analyses to 
multiple groups and multiple measures, (2) perform contrast analyses to compare
groups or measures, and (3) export and make basic changes to tables and figures.

## References
* Gurtman, M. B. (1992). Construct validity of interpersonal personality measures: The interpersonal circumplex as a nomological net. _Journal of Personality and Social Psychology, 63_(1), 105–118.

* Gurtman, M. B., & Pincus, A. L. (2003). The circumplex model: Methods and research applications. In J. A. Schinka & W. F. Velicer (Eds.), _Handbook of psychology. Volume 2: Research methods in psychology_ (pp. 407–428). Hoboken, NJ: John Wiley & Sons, Inc.

* Wright, A. G. C., Pincus, A. L., Conroy, D. E., & Hilsenroth, M. J. (2009). Integrating methods to optimize circumplex description and comparison of groups. _Journal of Personality Assessment, 91_(4), 311–322.

* Zimmermann, J., & Wright, A. G. C. (2017). Beyond description in interpersonal construct validation: Methodological advances in the circumplex Structural Summary Approach. _Assessment, 24_(1), 3–23.
